DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-10-12 Paper_May_2012/main.tex
Wie gestern besprochen, hier die Einleitung zum Kapitel wie man StatPDE loest.

%%% User's macros placed after this line %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\IfFileExists{dsfont.sty}%
%\usepackage{dsfont}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}

@@

\begin{abstract}
In this article, we describe the software library 
\textit{Deal.II Optimization Environment} (DOpElib).
The main feature of DOpElib is that it provides a unified interface to high level algorithms 
such as time stepping methods, nonlinear solvers and optimization routines. This structure ensures 
that first of all the user is required to write those sections of code that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with the need for just a few lines of code to change.
The article illustrates performance and features 
of the software package by four numerical tests. 
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
The \textit{Deal Optimization Environment} (DOpElib) 
provides a software toolkit to solve forward PDE
problems as well as optimal control problems constrained by a PDE. 
Its main feature is to give a unified interface to high level algorithms such as 
time stepping methods, nonlinear solvers and optimization routines. 
We aim that the user should only need to write those parts
of the code that are problem dependent while all invariant parts of the algorithms
should be reusable without any need for further coding.
In particular, the user should be able to switch between various different 
algorithms without the need to rewrite the problem dependent code, though he or she will
have to replace the algorithm object with an other one.

The
solution of a broad variety of PDE is possible in other software
libraries (like in deal.II \cite{deal}, dune \cite{dune} , 
or commercial solvers like Ansys \cite{ansys}) 
as well, but
DOpE concentrates on a unified approach for both linear and nonlinear
problems by interpreting every PDE problem as nonlinear and applying a
Newton method to solve it. While deal.II leaves much of the work and many
decisions to the user, DOpE intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimal control problems with PDE
constraints is an innovation in the DOpE framework.
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and fluid-structure interactions.

The DOpElib project is 
based on the \textit{deal.II} \cite{deal} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{deal}.
The authors acknowledge their past experience as well as discussions with 
the authors of the libraries 
Gascoigne/RoDoBo project, which was initiated by 
Roland Becker, Dominik Meidner,  and Boris Vexler \cite{rodobo}. 
From which some of the ideas to modularize the algorithms have arisen.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%The aim of DOpE is to provide a software toolkit to solve forward PDE
%problems as well as optimal control problems constrained by PDE. The
%solution of a broad variety of PDE is possible in deal.II as well, but
%DOpE concentrates on a unified approach for both linear and nonlinear
%problems by interpreting every PDE problem as nonlinear and applying a
%Newton method to solve it. While deal.II leaves much of the work and many
%decisions to the user, DOpE intends to be user-optimized by delivering
%prefabricated tools which require from the user only adjustments connected
%to his specific problem. The solution of optimal control problems with PDE
%constraints is an innovation in the DOpE framework.
%The focus is on the numerical solution of both stationary and nonstationary
%problems which come from different application fields, like elasticity and
%plasticity, fluid dynamics, and fluid-structure interactions.
%

At the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from deal.II including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Self written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Goal-oriented mesh adaptation with the dual-weighted residual method.
\item Different spatial triangulations for control and state variables.
\end{itemize}

@@

%The rest of this document is structured as follows: We start with an introduction in
%Chapter~\ref{chap:intro} where you will learn what is needed to run {\tt DOpElib}. 
%Further you will learn what problems we can solve and how all the different classes 
%work together for this purpose. This should help you figure out what the different classes
%do if you are in need of writing your own algorithm.
%
%Then assuming that you can work to your satisfaction with the algorithms already implemented
%we will show you how to create your own running example in Chapter~\ref{chap:howtoex}.
%This will be followed by a detailed description of all examples already shipped with 
%the library. You can find the examples for the solution of PDEs in Chapter~\ref{PDE}
%and those for the solution of optimization problems with PDEs in Chapter~\ref{OPT}.
%
%These notes conclude with a section that explains how we do automated testing of the 
%implementation in Chapter~\ref{chap:test}. This chapter will be of interest only if you 
%are trying to implement some new features to the library so that you can check that 
%the new code did not break anything.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
%\todo{Aus der Sicht des Users anhand eines konkreten Beispiels beschreiben}
%\todo{Einfaches Beispiel nehmen}
%\todo{erwaehnen, dass wir alles immer nicht-linear betrachten}
%\todo{Aussehen eines typischen PDE und Opt loop und deren Gemeinsamkeiten
%erklaeren. Vielleicht mir Pseudocode}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). 
\subsection{Solving of PDEs}
\subsubsection{Stationary problems}
The easiest case 
is that of a PDE in weak form to find some $u$
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$. As example, we take 
the Laplace problem in weak form:
\[
a(u)(\phi) = (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
To solve the problem, we need a configuration (a mesh), eventually 
a diffusion parameter, a finite element, a quadrature rule,
a linear solver (here the CG solver), and some output such as
a graphical solution and eventually some measurement functional.
Each of those routines are written in the \texttt{main.cc} file. 
We start by including necessary files both the DOpElib library and
the deal.II software:
\begin{lstlisting}
  #include "pdeproblemcontainer.h" etc. ...
\end{lstlisting}
In the \texttt{main} function, the following steps need to be done. First, 
the parameter reader for runtime parameters might be initialized. In a parameter
file the user might specifiy material paramters for the concrete PDE, but also
general parameters for nonlinear and linear solvers, and output-related 
aspects.

A linear finite element in two dimensions and two components is 
initialized:
\begin{lstlisting}
  FESystem<2> state_fe(FE_Q<2>(1), 2);
\end{lstlisting}
and a quadrature rule of order three is used to approximate the 
integrals:
\begin{lstlisting}
  QGauss<2> quadrature_formula(3);
\end{lstlisting}
Then, the \texttt{LocalPDE} object comes into play. This is an quite 
important feature of DOpElib because it allows for easy implementation 
of the variational formulation. Specific explication is given below because
the implementation is seperated in \texttt{localpde.h}.
Afterwards, a \texttt{LocalPointFunctional} is declared. As for the PDE, the 
implementation is done in another file, namely \texttt{localfunctional.h}.

Generically, each problem in DOpElib is considered to be time-dependent and nonlinear because
these are the 'real' world problems. Therefore, (even for the linear and stationary Laplace equation),
we have to declare a (pseudo) time:
\begin{lstlisting}
  std::vector<double> times(1, 0.);
\end{lstlisting}
Then, the spatial mesh (here the unit square) is created and three-times globally-refined:
\begin{lstlisting}
  GridGenerator::hyper_cube(triangulation, 0, 1);
  triangulation.refine_global(3);
\end{lstlisting}
A major component which organizes the whole solution process (independent from 
the problem!) is the 
\begin{lstlisting}
  MethodOfLines_StateSpaceTimeHandler<...> DOFH (...);
\end{lstlisting}
This method handles all dofs in space and time and different versions exist
for forward as well as optimization problems.

@@

  P.SetDirichletBoundaryColors(0, comp_mask, &DD1);
\end{lstlisting}
Next, the solver for the linear equations needs to be specified:
\begin{lstlisting}
  SSolver solver(&P, "fullmem", pr, idc);
\end{lstlisting}
Finally, the ouput handler is initialized via
\begin{lstlisting}
  P.RegisterOutputHandler(&out);
  solver.RegisterOutputHandler(&out);
\end{lstlisting}
Finally, the programm is started with:
\begin{lstlisting}
  solver.ReInit();
  out.ReInit();
  out.Write(outp, 1, 1, 1);
  solver.ComputeReducedFunctionals();
\end{lstlisting}
After explaining the basic compononents that are necessary (and always the same in DOpElib) 
for all  (linear and stationary) partial differential equations, we shall give 
an overview to the implementation of the specific form of the PDE. 
Beside specification of boundary conditions, the form of the local PDE depends 
on the concrete problem under consideration. 
This step is done in the \texttt{localpde.h} file. As already explained,
each PDE is treatet in a nonlinear manner such that for solving 
the Laplace problem the following nonlinear step is necessary to recapitulate:
Find $\delta u\in V$:
\[
(\nabla \delta u, \nabla \phi) = -(\nabla u, \nabla \phi) 
\]
and $u^{n+1} = u^n + \delta u$ for $n=1,2,3,ldots$.
Therefore, the implementation reads for the right hand side:
\begin{lstlisting}
   void CellEquation(...)
    {
      // implement (\nabla u, \nabla \phi) 
    }
\end{lstlisting}
Specifically, the inner loops reads
\begin{lstlisting}
   for (all quadrature points per cell q)
      {
        Tensor<2, 2> ugrads; // solution vector

@@

\subsubsection{Nonstationary problems}
In this section, we focus on the extension to 
the numerical solution of nonstationary processes. 
As example, we consider the heat equation:
Find $u$:
\[
\partial_t u - \Delta u = 0
\]
and in weak formulation:
Find $u$:
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$ and 
\[
a(u)(\phi) = (\partial_t u, \phi) + (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $u^n$, find $u^{n+1}$:
\[
\frac{(u^{n+1} - u^{n}, \phi)}{k} + \theta (\nabla u^{n+1}, \nabla \phi)
+ (1 - \theta) (\nabla u^{n}, \nabla \phi) = 0,
\]
Reordering in new and old terms and multiplication by the time step size $k$:
\[
(u^{n+1},\phi) + k \theta (\nabla u^{n+1}, \nabla \phi)
= (u^n, \phi) - (1 - \theta) (\nabla u^{n}, \nabla \phi).
\]
In DOpElib, different time stepping schemes can be found 
in the folder \texttt{DOpEsrc/tsschemes}.

@@

        local_cell_vector(i) +=  
          scale * (u * phi_i_u) * 
          state_fe_values.JxW(q_point);
      }
   }
\end{lstlisting}
Here, the advantage is that DOpElib orders the terms in the timestepping
problems
such that only $(u^{n+1},\phi)$ must be implemented. However, if someone 
needs explicit input of the old timestep term, then he can use 
\begin{lstlisting}
  void CellTimeEquationExplicit (...)
   {
     // implement (u^{n+1},\phi) and (u^n,\phi)  
   }
\end{lstlisting}
For instance, this explicit term is used for fluid-structure interaction in
which
the old timestep term needs explicit modification.

@@

\subsection{Independent features}
In this final section, we shall give a brief summary of features 
that are applicable (in general) to all types of problems. Specifically,
DOpElib provides goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method \cite{BeRa96}. Here, a certain 
cost functional such as a point evaluation, line intergration or domain
integration is considered. Determining the lowest error between the 
(unknown) exact solution and its numerical approximation for lowest computational
cost is the main goal of the DWR method.

Moreover, DOpElib provides standard error estimators based on residual evaluation and 
a ZZ error estimator, i.e., measureing the smoothness of the discrete first derivatives.
The latter one is accessed via deal.II.

In addition, DOpElib provides multi-mesh procedures 
\todo{kann Winni hier etwas schreiben?}

@@

%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\subsection{Interfaces to other software packages}
%Since DOpE uses as basis the software library deal.II, all its interfaces
%to other packages can be accessed to. For instance, such libararies 
%are trilinos for with an algebraic multigrid method, MPI for a parallel
%solution of the problem.
%
%In addition DOpE itself has an interface to SNOPT.
%
%%\section{Documentation, Code Development, Webpage}
%%\label{documentation}
%%At the present step, the DOpElib project comes with 
%%an detailed documentation of all features and examples 
%%in pdf format and an detailed programming code 
%%documentation. 
%%
%%\todo{License? How to access to DOpE?}
%%\todo{Name of the website}
%%
%%In addition, the DOpElib test suite provides some regression tests. 
%%They are run to compare 
%%the output to previous outputs. This is useful (necessary) after 
%%changing programming code anywhere in the library. If a test
%%succeeds, everything is fine in the library. If not, you should not
%%check in your code into DOpE. Please make sure what is going wrong and WHY!
%%Every command is computed via a Makefile. 
%%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}
\label{applications}
We present three numerical tests which demonstrate different
features of DOpElib:
\begin{itemize}
\item In the first numerical example, we consider 
goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method for the stationary Navier-Stokes equations.
\item The second example presents nonstationary fluid-structure 
interaction. The challenges for those kind of problems are the multi-domain
character of the coupled problem and the treatment of the coupling conditions.
The problem is formulated in a variational monolithically-coupled way which 
allows to consider goal-oriented mesh refinement and gradient-based optimization.
\item In the third numerical test, an optmization problem for structural mechanics
is discussed.
\end{itemize}

@@

Since the problem is fully nonstationary, the 
dynamics are presents in Figure \ref{res:fsi_2_mesh_and_x_velo}. 
By comparison of our findings in Figure \ref{res:results_ux_and_uy_fsi_2}
with the literature, it can be identified that 
the implementation in DOpElib is correct.

>>>>

%%% User's macros placed after this line %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\IfFileExists{dsfont.sty}%
%\usepackage{dsfont}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\deal}{\texttt{deal.II}}
\newcommand{\dope}{\texttt{DOpElib}}

@@

\begin{abstract}
In this article, we describe the software library 
\textit{Deal.II Optimization Environment} (\dope).
The main feature of \dope is that it provides a unified interface to high level algorithms 
such as time stepping methods, nonlinear solvers and optimization routines. This structure ensures 
that first of all the user is required to write those sections of code that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with the need for just a few lines of code to change.
The article illustrates performance and features 
of the software package by four numerical tests. 
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
The \textit{Deal Optimization Environment} (\dope) 
provides a software toolkit to solve forward PDE
problems as well as optimal control problems constrained by a PDE. 
Its main feature is to give a unified interface to high level algorithms such as 
time stepping methods, nonlinear solvers and optimization routines. 
We aim that the user should only need to write those parts
of the code that are problem dependent while all invariant parts of the algorithms
should be reusable without any need for further coding.
In particular, the user should be able to switch between various different 
algorithms without the need to rewrite the problem dependent code, though he or she will
have to replace the algorithm object with an other one.

The
solution of a broad variety of PDE is possible in other software
libraries (like in \deal \cite{deal}, dune \cite{dune} , 
or commercial solvers like Ansys \cite{ansys}) 
as well, but
\dope concentrates on a unified approach for both linear and nonlinear
problems by interpreting every PDE problem as nonlinear and applying a
Newton method to solve it. While \deal leaves much of the work and many
decisions to the user, \dope intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimal control problems with PDE
constraints is an innovation in the \dope framework.
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and fluid-structure interactions.

The \dope project is 
based on the \deal \cite{deal} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{deal}.
The authors acknowledge their past experience as well as discussions with 
the authors of the libraries 
Gascoigne/RoDoBo project, which was initiated by 
Roland Becker, Dominik Meidner,  and Boris Vexler \cite{rodobo}. 
From which some of the ideas to modularize the algorithms have arisen.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%The aim of \dope is to provide a software toolkit to solve forward PDE
%problems as well as optimal control problems constrained by PDE. The
%solution of a broad variety of PDE is possible in \deal as well, but
%\dope concentrates on a unified approach for both linear and nonlinear
%problems by interpreting every PDE problem as nonlinear and applying a
%Newton method to solve it. While \deal leaves much of the work and many
%decisions to the user, \dope intends to be user-optimized by delivering
%prefabricated tools which require from the user only adjustments connected
%to his specific problem. The solution of optimal control problems with PDE
%constraints is an innovation in the \dope framework.
%The focus is on the numerical solution of both stationary and nonstationary
%problems which come from different application fields, like elasticity and
%plasticity, fluid dynamics, and fluid-structure interactions.
%

At the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from \deal including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Self written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Goal-oriented mesh adaptation with the dual-weighted residual method.
\item Different spatial triangulations for control and state variables.
\end{itemize}

@@

%The rest of this document is structured as follows: We start with an introduction in
%Chapter~\ref{chap:intro} where you will learn what is needed to run {\tt \dope}. 
%Further you will learn what problems we can solve and how all the different classes 
%work together for this purpose. This should help you figure out what the different classes
%do if you are in need of writing your own algorithm.
%
%Then assuming that you can work to your satisfaction with the algorithms already implemented
%we will show you how to create your own running example in Chapter~\ref{chap:howtoex}.
%This will be followed by a detailed description of all examples already shipped with 
%the library. You can find the examples for the solution of PDEs in Chapter~\ref{PDE}
%and those for the solution of optimization problems with PDEs in Chapter~\ref{OPT}.
%
%These notes conclude with a section that explains how we do automated testing of the 
%implementation in Chapter~\ref{chap:test}. This chapter will be of interest only if you 
%are trying to implement some new features to the library so that you can check that 
%the new code did not break anything.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
%\todo{Aus der Sicht des Users anhand eines konkreten Beispiels beschreiben}
%\todo{Einfaches Beispiel nehmen}
%\todo{erwaehnen, dass wir alles immer nicht-linear betrachten}
%\todo{Aussehen eines typischen PDE und Opt loop und deren Gemeinsamkeiten
%erklaeren. Vielleicht mir Pseudocode}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). 
In particular, we remove all tedious routine implementation needed in 
\deal by introducing appropriate classes that take care of all the 
problem independent implementation while leaving only the problem specific 
part to the user. To show how this is designed we will, in the sequel, 
describe the reasoning first for a simple stationary PDE and then show
how this idea extends to non stationary problems as well as to optimization
problems with PDE constraints.

\subsection{Solving of PDEs}
\subsubsection{Stationary problems}
For the start we consider a prototypical problem given by the solution of 
a stationary PDE, which we assume to be given in weak form.
In abstract notation it reads; find some $u$ in an appropriate set 
such that
\[
a(u)(\phi) = f(\phi) \quad \forall \phi \in V,
\]
with some appropriate space $V$ for the test functions and a 
right hand side $f$.
To illustrate this further, we take the stationary 
incompressible Navier-Stokes equations in a domain $\Omega \subset \R^d$, 
i.e., $u = (v,p)$ and $\phi = (\phi_u,\phi_p)$ then
\[
a(u)(\phi) = (\nabla v, \nabla\phi_u) + (v \nabla v,\phi_u) + (p, \nabla \cdot \phi_u) + (\nabla \cdot v ,\phi_p)
\quad \forall \phi \in V = H^1_0(\Omega;\R^d) \times L^2_0(\Omega)
\]
Then the solution will be searched for in $V + (u_d,0)$ where $u_d$ is some 
continuation of the given Dirichlet boundary data.
Then typically we do not only want to calculate $u$ but more likely we are 
interested in one (or several) numbers $J(u)$.

In order to solve this problem, i.e., calculate $J(u)$, we need to discretize 
the equation, here using finite elements. This means we need to pick a 
subdivision $\mathcal T_h$, for instance into quadrilaterals,
of the domain $\Omega$ and a finite element, in the concrete case we could, 
for instance, take the $\mathcal Q_2/\mathcal Q_1$ Taylor-Hood element.
As it is clear, that the initialization of the unknowns, the local shape 
functions, removal of hanging node constraints, etc. always follows the 
same procedure, we define an object which takes the finite element (FE)
and the triangulation and takes care of all other things related with 
handling the finite element space $V_h$. We will refer, to 
this as the \texttt{SpaceTimeHandler}. It is then clear, that the simplest 
form of the constructor should have the following interface:
\begin{lstlisting}
  template<typename Triangulation, typename FE>
      void SpaceTimeHandler(Triangulation &, FE);
\end{lstlisting}
Next, one needs to solve the discrete equation
\[
a(u_h)(\phi_h) = f(\phi_h) \quad \forall \phi \in V_h.
\]
To this end one can employ a newton type method with some globalization 
like line-search. To do so, one needs to calculate the residual of the 
equation, as well as solving linearized equations for the calculation of 
the newton method and then employing some kind of globalization strategy.
This again will usually require the evaluation of the residual.

To evaluate the residual will require the calculation of the integrals 
involved in the weak formulation of the problem. For example 
the $i$-th component of the residual, with respect to the given FE-basis 
$\phi_i$, will read as follows
\[
 a(u_h)(\phi_i) - f(\phi_i) = \sum_{K\in \mathcal T_h} \int_K \nabla u, \nabla \phi_i\,dx + \ldots
\]
Again, we see that, except for the local element integrals, all 
calculations are independent from the concrete form of the PDE. 
Hence we define an object \texttt{Integrator}
that offers a method to calculate the residual given a user-defined 
object \texttt{PDE} that implements the local element integrals, i.e.,
\begin{lstlisting}
  template<typename PDE,typename VECTOR>
      void Integrator
             ::ComputeNonlinearResidual(PROBLEM& pde, 
                                        VECTOR & residual);
\end{lstlisting}
where \texttt{VECTOR} is one of the possible vector types given by \deal.
It is clear that we need to ensure that the integrator object has access
to the information stored in the \texttt{SpaceTimeHandler}
to correctly loop over all elements.
To solve the linearized systems, again, all that is left is the calculation
of the matrix corresponding to the linearized equation. To calculate this,
again we can define a method in the integrator that does exactly this provided
we have a user defined object \texttt{PDE} that can evaluate the local 
integrals needed for assembling the matrix, i.e.,
\begin{lstlisting}
  template<typename PDE,typename MATRIX>
      void Integrator
             ::ComputeMatrix(PROBLEM& pde, 
                             MATRIX & system_matrix);
\end{lstlisting}
where now \texttt{MATRIX} is one of the matrix types supported by \deal.
With this we can now define a nonlinear solver \texttt{NewtonSolver} 
for the solution of the discrete problem requiring the existence of the 
following function
\begin{lstlisting}
  template<typename PDE, typename VECTOR>
      void NewtonSolver
             ::NonlinearSolve(PROBLEM& pde, 
                              VECTOR & solution);
\end{lstlisting}

Finally, we note that to compute the requested functional value 
all that is left is to evaluate the functional in the computed solution 
which can be done again using an appropriate integrator method.
Since the sequence of events sketched above is again independent of the 
exact problem we introduce another container around it which we call 
\texttt{StatPDEProblem} that then simply provides the user with the 
method
\begin{lstlisting}
      void StatPDEProblem::ComputeReducedFunctionals();
\end{lstlisting}
That combines all the necessary steps to calculate $J(u_h)$.

\begin{remark}
In fact, the above constructors are not exactly what is implemented in \dope.
This is due to the fact, that for instance 
there are several different sparsity patterns available in 
\deal which the user should be able chose. Thus it needs to be 
given to the \texttt{SpaceTimeHandler}. However, to keep the 
presentation clear we refrained from including these details above and in
what follows.
\end{remark}

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}

To solve the problem, we need a configuration (a mesh), eventually 
a diffusion parameter, a finite element, a quadrature rule,
a linear solver (here the CG solver), and some output such as
a graphical solution and eventually some measurement functional.
Each of those routines are written in the \texttt{main.cc} file. 
We start by including necessary files both the \dope library and
the \deal software:
\begin{lstlisting}
  #include "pdeproblemcontainer.h" etc. ...
\end{lstlisting}
In the \texttt{main} function, the following steps need to be done. First, 
the parameter reader for runtime parameters might be initialized. In a parameter
file the user might specify material parameters for the concrete PDE, but also
general parameters for nonlinear and linear solvers, and output-related 
aspects.

A linear finite element in two dimensions and two components is 
initialized:
\begin{lstlisting}
  FESystem<2> state_fe(FE_Q<2>(1), 2);
\end{lstlisting}
and a quadrature rule of order three is used to approximate the 
integrals:
\begin{lstlisting}
  QGauss<2> quadrature_formula(3);
\end{lstlisting}
Then, the \texttt{LocalPDE} object comes into play. This is an quite 
important feature of \dope because it allows for easy implementation 
of the variational formulation. Specific explication is given below because
the implementation is separated in \texttt{localpde.h}.
Afterwards, a \texttt{LocalPointFunctional} is declared. As for the PDE, the 
implementation is done in another file, namely \texttt{localfunctional.h}.

Generically, each problem in \dope is considered to be time-dependent and nonlinear because
these are the 'real' world problems. Therefore, (even for the linear and stationary Laplace equation),
we have to declare a (pseudo) time:
\begin{lstlisting}
  std::vector<double> times(1, 0.);
\end{lstlisting}
Then, the spatial mesh (here the unit square) is created and three-times globally-refined:
\begin{lstlisting}
  GridGenerator::hyper_cube(triangulation, 0, 1);
  triangulation.refine_global(3);
\end{lstlisting}
A major component which organizes the whole solution process (independent from 
the problem!) is the 
\begin{lstlisting}
  MethodOfLines_StateSpaceTimeHandler<...> DOFH (...);
\end{lstlisting}
This method handles all dofs in space and time and different versions exist
for forward as well as optimization problems.

@@

  P.SetDirichletBoundaryColors(0, comp_mask, &DD1);
\end{lstlisting}
Next, the solver for the linear equations needs to be specified:
\begin{lstlisting}
  SSolver solver(&P, "fullmem", pr, idc);
\end{lstlisting}
Finally, the output handler is initialized via
\begin{lstlisting}
  P.RegisterOutputHandler(&out);
  solver.RegisterOutputHandler(&out);
\end{lstlisting}
Finally, the programm is started with:
\begin{lstlisting}
  solver.ReInit();
  out.ReInit();
  out.Write(outp, 1, 1, 1);
  solver.ComputeReducedFunctionals();
\end{lstlisting}
After explaining the basic compononents that are necessary (and always the same in \dope) 
for all  (linear and stationary) partial differential equations, we shall give 
an overview to the implementation of the specific form of the PDE. 
Beside specification of boundary conditions, the form of the local PDE depends 
on the concrete problem under consideration. 
This step is done in the \texttt{localpde.h} file. As already explained,
each PDE is treatet in a nonlinear manner such that for solving 
the Laplace problem the following nonlinear step is necessary to recapitulate:
Find $\delta u\in V$:
\[
(\nabla \delta u, \nabla \phi) = -(\nabla u, \nabla \phi) 
\]
and $u^{n+1} = u^n + \delta u$ for $n=1,2,3,ldots$.
Therefore, the implementation reads for the right hand side:
\begin{lstlisting}
   void CellEquation(...)
    {
      // implement (\nabla u, \nabla \phi) 
    }
\end{lstlisting}
Specifically, the inner loops reads
\begin{lstlisting}
   for (all quadrature points per cell q)
      {
        Tensor<2, 2> ugrads; // solution vector

@@

\subsubsection{Nonstationary problems}
In this section, we focus on the extension to 
the numerical solution of nonstationary processes. 
As example, we consider the heat equation:
Find $u$:
\[
\partial_t u - \Delta u = 0
\]
and in weak formulation:
Find $u$:
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$ and 
\[
a(u)(\phi) = (\partial_t u, \phi) + (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $u^n$, find $u^{n+1}$:
\[
\frac{(u^{n+1} - u^{n}, \phi)}{k} + \theta (\nabla u^{n+1}, \nabla \phi)
+ (1 - \theta) (\nabla u^{n}, \nabla \phi) = 0,
\]
Reordering in new and old terms and multiplication by the time step size $k$:
\[
(u^{n+1},\phi) + k \theta (\nabla u^{n+1}, \nabla \phi)
= (u^n, \phi) - (1 - \theta) (\nabla u^{n}, \nabla \phi).
\]
In \dope, different time stepping schemes can be found 
in the folder \texttt{DOpEsrc/tsschemes}.

@@

        local_cell_vector(i) +=  
          scale * (u * phi_i_u) * 
          state_fe_values.JxW(q_point);
      }
   }
\end{lstlisting}
Here, the advantage is that \dope orders the terms in the timestepping
problems
such that only $(u^{n+1},\phi)$ must be implemented. However, if someone 
needs explicit input of the old timestep term, then he can use 
\begin{lstlisting}
  void CellTimeEquationExplicit (...)
   {
     // implement (u^{n+1},\phi) and (u^n,\phi)  
   }
\end{lstlisting}
For instance, this explicit term is used for fluid-structure interaction in
which
the old timestep term needs explicit modification.

@@

\subsection{Independent features}
In this final section, we shall give a brief summary of features 
that are applicable (in general) to all types of problems. Specifically,
\dope provides goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method \cite{BeRa96}. Here, a certain 
cost functional such as a point evaluation, line intergration or domain
integration is considered. Determining the lowest error between the 
(unknown) exact solution and its numerical approximation for lowest computational
cost is the main goal of the DWR method.

Moreover, \dope provides standard error estimators based on residual evaluation and 
a ZZ error estimator, i.e., measureing the smoothness of the discrete first derivatives.
The latter one is accessed via \deal.

In addition, \dope provides multi-mesh procedures 
\todo{kann Winni hier etwas schreiben?}

@@

%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\subsection{Interfaces to other software packages}
%Since DOpE uses as basis the software library \deal, all its interfaces
%to other packages can be accessed to. For instance, such libararies 
%are trilinos for with an algebraic multigrid method, MPI for a parallel
%solution of the problem.
%
%In addition DOpE itself has an interface to SNOPT.
%
%%\section{Documentation, Code Development, Webpage}
%%\label{documentation}
%%At the present step, the \dope project comes with 
%%an detailed documentation of all features and examples 
%%in pdf format and an detailed programming code 
%%documentation. 
%%
%%\todo{License? How to access to DOpE?}
%%\todo{Name of the website}
%%
%%In addition, the \dope test suite provides some regression tests. 
%%They are run to compare 
%%the output to previous outputs. This is useful (necessary) after 
%%changing programming code anywhere in the library. If a test
%%succeeds, everything is fine in the library. If not, you should not
%%check in your code into DOpE. Please make sure what is going wrong and WHY!
%%Every command is computed via a Makefile. 
%%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}
\label{applications}
We present three numerical tests which demonstrate different
features of \dope:
\begin{itemize}
\item In the first numerical example, we consider 
goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method for the stationary Navier-Stokes equations.
\item The second example presents nonstationary fluid-structure 
interaction. The challenges for those kind of problems are the multi-domain
character of the coupled problem and the treatment of the coupling conditions.
The problem is formulated in a variational monolithically-coupled way which 
allows to consider goal-oriented mesh refinement and gradient-based optimization.
\item In the third numerical test, an optmization problem for structural mechanics
is discussed.
\end{itemize}

@@

Since the problem is fully nonstationary, the 
dynamics are presents in Figure \ref{res:fsi_2_mesh_and_x_velo}. 
By comparison of our findings in Figure \ref{res:results_ux_and_uy_fsi_2}
with the literature, it can be identified that 
the implementation in \dope is correct.
